Growth curve analysis | data: 08/17/2013
========================================================
This script does some data plotting and quality check
for my first Bioscreen-C experiment  
hebin  
17 aug 2013  

Summary
------------------
1. The rich media condition is abnormal -- the first column on plate A didn't grow / _pho4_ (CG3) grows the best under this condition / the saturation yield of _pho4_ is the highest among all experimental conditions
2. 

Read data
------------------
Notes:  
* change the working directory
* replace the filename with the appropriate name
```{r Read bioscreen data, cache=TRUE}
raw <- read.csv("BioscreenExperiment20130816.csv")
# note: each column is temporal data for one well
data <- as.matrix(raw[,-1])
```

Design matrix
------------------
* define the strains
* define all factors involved
* setup the design matrix
```{r Create design matrix,echo=FALSE}
# --------------------------
# create the design matrix
# 1. change the number of strains 
#    and conditions here
n.strain = 20
n.condition = 10
# 2. name the strains used, including biological replicates
ST <- c("wt", "pho80-1", "pho80-2", "pho4", "pho80.scPho4-1", "pho80.scPho4-2", "empty")
# 3. give the order of strains, including technical replicates
strain <- rep( c( rep(ST[1:3],each=3), ST[7], 
                  rep(ST[4:6],each=3), ST[7] ), n.condition )
# 4. name the factors
factors <- c("Glucose","Pi","Rapamycin")
# 5. give the conditions for one strain, then replicate for all
condition <- list("Glucose"=factor(c(rep("2%",4), rep("0.5%",3), rep("2%",3)), levels=c("2%","0.5%")),
                  "Pi"=factor(paste(c(10,1,0.1,0,10,1,0.1,10,10,1),"mM",sep="")),
                  "Rapamycin"=factor(paste(c(rep(0,7),10,50,10),"ng/ml",sep="")))
conditions <- data.frame(condition) [ rep(seq_len(n.condition),each=n.strain), ]
# 6. finally, synthesize into a design matrix
design.mat <- data.frame(strain, conditions)
# 7. Optionally, give the time points
Time <- seq(0,24,0.25); attr(Time, "unit") <- "hr"
cat("Strains\n")
print(ST)
cat("Each strain is tested in the following 10 conditions\n")
print(unique(conditions))
```

Functions
-----------------
1. Subfunction to summarize, extract data
`myMeanSD <- function(x, d)`  
`extract1Col <- function(x, d, cond, strain, index, range=c(1:10))`
```{r Sub1, echo=FALSE}
myMeanSD <- function(x, d){
  # Purpose ==
  #   summarize the data by calculating the mean and sd of tech. repl.
  # Input ==
  #   x: data matrix
  #   d: design matrix
  # Output ==
  #   A list consisting of two arrays: mean and sd
  
  # combining the different conditions into a single factor
  d.tmp <- apply(d, 1, function(y) paste(y, collapse="  "))
  d.as.factor <- factor(d.tmp, levels=unique(d.tmp), ordered=TRUE)
  # calculate mean and sd
  mean <- apply(x,1,function(z) tapply(z,d.as.factor,mean))
  sd   <- apply(x,1,function(z) tapply(z,d.as.factor,sd  ))
  # recover the design matrix
  tmp  <- rownames(mean) # a vector of strings separated by "|"
  new.d <- sapply(tmp, function(z) strsplit(z,"  ")[[1]])
  new.d <- data.frame(t(new.d))
  colnames(new.d) <- names(d); rownames(new.d) <- NULL
  return(list("mean"=mean,"sd"=sd,"design.mat"=new.d))
}
extractCol <- function(x, d, strain, cond, index=1, range=c(2:25)){
  # Input ===
  #   x: raw data matrix
  #   d: design matrix
  #   strain: asis
  #   cond: a number as the index of the condition in the list "condition"
  #   index: index for which technical replicate to return, e.g. 1/2/3

  choose = d$strain %in% strain
  # limit to the designated conditions
  # assume factors and condition are global variables
  for ( z in factors ){
    tmp = d[,z] %in% condition[[z]][cond]
    choose = choose & tmp
  }
  x.tmp = x[range,choose]
  x.return = data.frame(od=x.tmp[,index], t=Time[range])
  return( x.return )
}
```

2. Simple plotting function
`simplot <- function(ex, mode="basic")`
```{r Sub1.1, echo=FALSE}
simplot <- function(ex, mode="basic"){
  # Input ==
  #   ex: a dataframe of extracted OD values and the corresponding time points
  #   mode: "basic"-simple x-y plot; "log"-y on log scale; "derivative"-first derivative using adjacent pair of values
  if( class(mode) == "character" ){
    mode = as.numeric(factor(mode, levels=c("basic","log","derivative")))
  }
  if( is.na(mode) ){
    warning("Mode should be basic/log/derivative")
  }
  if( mode == 2 | mode == 3 ){
    x = log2(ex$od/ex$od[1])
  }
  # basic
  if( mode == 1 ){
    plot(ex$t, ex$od, xlab=paste("Time", attr(ex$t,"unit")), ylab="OD600")
  }
  else if( mode == 2 ){
    plot(ex$t, x, xlab=paste("Time", attr(ex$t,"unit")), ylab="log2( normalized OD )")
  }
  # derivative
  else if( mode == 3 ){
    x.diff = x[-1] - x[-length(x)]
    plot(ex$t[-1], x.diff, xlab=paste("Time", attr(ex$t,"unit")), ylab="d_Log2(OD600)")
  }
}
```

2. A function to visualize the time course for any given strain at a particular condition  
`plotTrace <- function(x, d, strain, cond, ylim=NA, lwd=1)`  
```{r Sub2,echo=FALSE}
plotTrace <- function(x, d, strain, cond, ylim=NA, lwd=1){
# Input ==
#   x: data matrix
#   d: design matrix
#   strain: strain id, or "all"
#   cond: a vector of conditions to look at
# Output ==
#   Line plot
  
  # limit to the strain(s)
  if(strain[1] == "all"){
    strain = levels(d$strain)
  }
  choose = d$strain %in% strain
  # limit to the designated conditions
  # assume factors and condition are global variables
  for ( z in factors ){
    tmp = d[,z] %in% condition[[z]][cond]
    choose = choose & tmp
  }
  # prepare data for plot
  x.plot <- x[,choose]
  d.plot <- d[choose,]
  d.tmp <- apply(d.plot[,-1], 1, function(y) paste(y, collapse=" "))
  d.as.factor <- factor(d.tmp, levels=unique(d.tmp), ordered=TRUE)
  strain.as.factor = factor(d.plot[,1], levels=unique(d.plot[,1], ordered=TRUE))
  if( is.na(ylim) ){
    ylim = range(x.plot)
  }
  thin <- seq.int(1L,nrow(x),20L)
  # change layout
  layout(matrix(1:3,ncol=1),heights=c(6,1,1))
  # plot main figure
  oldpar <- par(mar=c(5,5,2,2))
  plot(Time, x.plot[,1], type="l", ylim=ylim, lwd=lwd, col=as.numeric(d.as.factor[1]), 
       xlab="Time (hr)", ylab="OD600",cex.lab=1.3, cex.axis=1.2, main="Growth curve")
  points(Time[thin], x.plot[thin,1], pch=as.numeric(strain.as.factor[1]),cex=1.2,
         col=as.numeric(d.as.factor[1]))
  for ( j in 2:ncol(x.plot) ){
    lines(Time, x.plot[,j], ylim=ylim, lwd=lwd, col=rep(1:6,3)[as.numeric(d.as.factor[j])], 
          lty=rep(1:3,each=6)[as.numeric(d.as.factor[j])]) 
    points(Time[thin], x.plot[thin,j], pch=as.numeric(strain.as.factor[j]),cex=1.2,
           col=rep(1:6,3)[as.numeric(d.as.factor[j])] )
  }
  # plot legend for condition
  par(mar=rep(0,4))
  plot.new()
  legend("center","groups",legend=levels(d.as.factor),cex=1.3,bty="n",
         col=rep(1:6,3)[1:length(unique(d.as.factor))],
         title=paste(names(d.plot)[-1],collapse="/"),
         ncol=min( 4, length( unique(d.as.factor) ) ),
         lty=rep(1:3,each=6)[1:length(unique(d.as.factor))] )
  # plot legend for strain
  plot.new()
  legend("center","groups",legend=levels(strain.as.factor),lty=1,cex=1.3,bty="n",
         pch=1:length(unique(strain.as.factor)),
         title="Strains",ncol=min(4,length(unique(strain.as.factor))))
  par(oldpar)
}
```

3. Based on plotTrace, but plot the mean of tech. repl.  
`plotMeanTrace <- function(X, strain, cond, title="", ylim=NA, lwd=1)`  
`plainMeanTrace <- function(X, strain, cond, title="", ylim=NA, lwd=1)`  
  _Same as plotMeanTrace, except it doesn't try to plot the legend_
```{r Sub3,echo=FALSE}
plotMeanTrace <- function(X, strain, cond, title="", ylim=NA, lwd=1){
# Purpose ==
#   Similar to plotTrace, but use summary of tech. repl.
# Input ==
#   X: output of myMeanSD, contains three elements -- mean, sd, (new)design.mat
#   strain: strain id, or "all"
#   cond: a vector of conditions to look at
# Output ==
#   Line plot
# Note ==
#   Now configured to plot the mean of the tech. repl.
  
  # limit to the strain(s)
  mean = t(X$mean); sd = t(X$sd); d = X$design.mat
  if(strain[1] == "all"){
    strain = levels(d$strain)
  }
  choose = d$strain %in% strain
  # limit to the designated conditions
  # assume factors and condition are global variables
  for ( z in factors ){
    tmp = d[,z] %in% condition[[z]][cond]
    choose = choose & tmp
  }
  # prepare data for plot
  mean.plot <- mean[,choose]
  sd.plot <- sd[,choose]
  d.plot <- d[choose,]
  d.tmp <- apply(d.plot[,-1], 1, function(y) paste(y, collapse=" "))
  d.as.factor <- factor(d.tmp, levels=unique(d.tmp), ordered=TRUE)
  strain.as.factor = factor(d.plot[,1])
  if( is.na(ylim[1]) ){
    ylim = range(mean.plot)
  }
  thin <- seq.int(1L,nrow(mean.plot),20L)
  # change layout
  layout(matrix(1:3,ncol=1),heights=c(6,1,1))
  # plot main figure
  oldpar <- par(mar=c(5,5,2,2))
  plot(Time, mean.plot[,1], type="l", ylim=ylim, lwd=lwd, col=as.numeric(d.as.factor[1]),
       xlab="Time (hr)", ylab="OD600",cex.lab=1.3, cex.axis=1.2, main=title, cex.main=1.5)
  points(Time[thin], mean.plot[thin,1], pch=as.numeric(strain.as.factor[1]),cex=1.2,
         col=as.numeric(d.as.factor[1]))
  arrows(Time[thin], mean.plot[thin,1]-sd.plot[thin,1], Time[thin], mean.plot[thin,1]+sd.plot[thin,1], 
         length=0.02, angle=90, code=3, lwd=0.5,
         col=as.numeric(d.as.factor[1]))
  for ( j in 2:ncol(mean.plot) ){
    lines(Time, mean.plot[,j], ylim=ylim, lwd=lwd, col=rep(1:6,3)[as.numeric(d.as.factor[j])], 
          lty=rep(1:3,each=6)[as.numeric(d.as.factor[j])]) 
    points(Time[thin], mean.plot[thin,j], pch=as.numeric(strain.as.factor[j]),cex=1.2,
           col=rep(1:6,3)[as.numeric(d.as.factor[j])] )
    arrows(Time[thin], mean.plot[thin,j]-sd.plot[thin,j], Time[thin], mean.plot[thin,j]+sd.plot[thin,j], 
         length=0.02, angle=90, code=3, lwd=0.5, 
         col=as.numeric(d.as.factor[j]))
  
  }
  # plot legend for condition
  par(mar=c(0,0,0,0))
  plot.new()
  legend("center","groups",legend=levels(d.as.factor),cex=1.3,bg=rgb(1,1,0,0.3),
         col=rep(1:6,3)[1:length(unique(d.as.factor))],
         title=paste(names(d.plot)[-1],collapse="/"),
         ncol=min( 4, length( unique(d.as.factor) ) ),
         lty=rep(1:3,each=6)[1:length(unique(d.as.factor))] )
  # plot legend for strain
  plot.new()
  legend("center","groups",legend=levels(strain.as.factor),lty=1,cex=1.3,bg=rgb(1,1,0,0.3),
         pch=1:length(unique(strain.as.factor)),
         title="Strains",ncol=min(4,length(unique(strain.as.factor))))
  par(oldpar)
}
plainMeanTrace <- function(X, strain, cond, title="", ylim=NA, lwd=1){
# Purpose ==
#   Same as plotMeanTrace, except it doesn't try to plot the legend
# Input ==
#   X: output of myMeanSD, contains three elements -- mean, sd, (new)design.mat
#   strain: strain id, or "all"
#   cond: a vector of conditions to look at
# Output ==
#   Line plot
# Note ==
#   Now configured to plot the mean of the tech. repl.
  
  # limit to the strain(s)
  mean = t(X$mean); sd = t(X$sd); d = X$design.mat
  if(strain[1] == "all"){
    strain = levels(d$strain)
  }
  choose = d$strain %in% strain
  # limit to the designated conditions
  # assume factors and condition are global variables
  for ( z in factors ){
    tmp = d[,z] %in% condition[[z]][cond]
    choose = choose & tmp
  }
  # prepare data for plot
  mean.plot <- mean[,choose]
  sd.plot <- sd[,choose]
  d.plot <- d[choose,]
  d.tmp <- apply(d.plot[,-1], 1, function(y) paste(y, collapse=" "))
  d.as.factor <- factor(d.tmp, levels=unique(d.tmp), ordered=TRUE)
  strain.as.factor = factor(d.plot[,1])
  if( is.na(ylim[1]) ){
    ylim = range(mean.plot)
  }
  thin <- seq.int(1L,nrow(mean.plot),20L)
  # plot main figure
  oldpar <- par(mar=c(5,5,2,2))
  plot(Time, mean.plot[,1], type="l", ylim=ylim, lwd=lwd, col=as.numeric(d.as.factor[1]),
       xlab="Time (hr)", ylab="OD600",cex.lab=1.3, cex.axis=1.2, main=title, cex.main=1.5)
  points(Time[thin], mean.plot[thin,1], pch=as.numeric(strain.as.factor[1]),cex=1.2,
         col=as.numeric(d.as.factor[1]))
  arrows(Time[thin], mean.plot[thin,1]-sd.plot[thin,1], Time[thin], mean.plot[thin,1]+sd.plot[thin,1], 
         length=0.02, angle=90, code=3, lwd=0.5, 
         col=as.numeric(d.as.factor[1]))
  for ( j in 2:ncol(mean.plot) ){
    lines(Time, mean.plot[,j], ylim=ylim, lwd=lwd, col=rep(1:6,3)[as.numeric(d.as.factor[j])], 
          lty=rep(1:3,each=6)[as.numeric(d.as.factor[j])]) 
    points(Time[thin], mean.plot[thin,j], pch=as.numeric(strain.as.factor[j]),cex=1.2,
           col=rep(1:6,3)[as.numeric(d.as.factor[j])] )
    arrows(Time[thin], mean.plot[thin,j]-sd.plot[thin,j], Time[thin], mean.plot[thin,j]+sd.plot[thin,j], 
         length=0.02, angle=90, code=3, lwd=0.5, 
         col=as.numeric(d.as.factor[j]))
  }
  # plot legend for conditons
  legend("topleft",legend=levels(d.as.factor),
         col=rep(1:6,3)[1:length(unique(d.as.factor))],
         title=paste(names(d.plot)[-1],collapse="/"),
         lty=rep(1:3,each=6)[1:length(unique(d.as.factor))] )
  # plot legend for strain
  legend("bottomright",legend=levels(strain.as.factor),lty=1,
         pch=1:length(unique(strain.as.factor)),
         title="Strains")
  par(oldpar)
}
```

4. Estimate growth rate
```{r Sub4, echo=FALSE}
growthRate <- function(ex, mode=1, plot=TRUE){
  # Purpose ==
  #   Estimate the doubling time from growth curve
  # Input ==
  #   ex: data frame, with od and time
  #   mode: 1 or 2, decide the type of diagnostic plot to output
  # Output ==
  #   0. return two stats: growth rate and yield
  #   1. plot either the log2(OD)~Time and show the linear fit (mode=1) or plot the Growth rate in units of min per doubling by calculating the slope based on any adjacent pair of points (mode=2)
  #   2. If mode=1, return the doubling time in unit of min
  ex.log = log2(ex$od/ex$od[1]); time = ex$t
  if( mode == 1 ){
    fit <- lm(ex.log~time)
    if( plot ){
      simplot(ex, mode=2)
      abline(fit, col="red", lwd=.5)
      legend("topleft",bty="n",
             legend=gettextf("y = %.2f + %.2f x",coef(fit)[1], coef(fit)[2]))
    }
    doubling.time <- 1/coef(fit)[2]*60
  }
  else if( mode == 2 ){
    if( plot ){ simplot(ex, mode=3) }
    ex.diff = ex.log[-1]-ex.log[-length(ex.log)]
    t.diff = time[-1] - time[-length(time)]
    # find out the maximum rate, or minimum doubling time
    if( sum(ex.diff<0) >= 8 ){
      min.time <- -1000
    }
    else{
      i.min <- which.max(ex.diff)
      min.time <- t.diff[i.min] / ex.diff[i.min]
    }
    # return the minimum doubling time
    doubling.time <- min.time * 60
  }
  return( "doubling time" = doubling.time )
}
```

Analyse
-----------------
### 1. Quality check
### _Negative controls_
The last well of each column has no cells in it. Plotting the OD of the 20 empty wells should be background.
Notice that rapamycin adds to the background. My current analysis didn't remove the background.
```{r negative controls, fig.width=10, fig.height=7.5, warning=FALSE}
x = myMeanSD(data, design.mat)
plotMeanTrace(x, "empty", 1:10, title="Empty wells", lwd=1.5)
```

### _But something is wrong with the first column_
Notice the flat line including the wt, the two pho80 strain, all of which are in the first column
Also notice that the rest three strains shared a similar pattern, i.e. there is a peak at ~6h, then a drop, followed by a recovery. In the case of pho80.scPHO4, there is a subsequent drop. The _pho4_ strain seems to do particularly well in this condition. By the end it still doesn't seem to saturate.  
```{r first column problem, warning=FALSE}
plotMeanTrace(x, "all", 1, lwd=1.5, title="First column problem")
```
__Questions 1.__ Why is there a drop in OD at ~6hr, what is limiting?
_Calculate the rate of Pi consumption in rich media_
My estimate 


### 2. Overview and comparisons between strains / conditions
### _Starved for Pi_
The biological replicates (pho80 and pho80.scPHO4) are quite similar. From now on I just plot one of them
```{r Pi starvation, warning=FALSE, fig.width=12, fig.height=6}
st.use <- c("wt","pho80-1","pho4","pho80.scPho4-1")
layout(matrix(c(1,2),ncol=2))
plainMeanTrace(x, st.use, c(2:3), lwd=1.5, title="Low Pi")
plainMeanTrace(x, st.use, 4, lwd=1.5, title="No Pi")
```

### _Starved for Glucose (and Pi)_
```{r Glucose and Pi starvation, warning=FALSE, fig.width=12,fig.height=6}
layout(matrix(c(1,2),ncol=2))
plainMeanTrace(x, st.use, 5, lwd=1.5, ylim=c(0.06,1.2), title="Low Glucose")
plainMeanTrace(x, st.use, c(6,7), lwd=1.5, ylim=c(0.06,1.2), title="Low Glucose + Low Pi")
```

### _Comparison grouped by strains, not conditions_
```{r Grouped by strains, warning=FALSE, fig.width=12, fig.height=12}
layout( matrix(c(1:4), byrow=TRUE, ncol=2) )
plainMeanTrace(x, "wt", 1:7, lwd=1.5, ylim=c(0.06,1.2), title="wt")
plainMeanTrace(x, "pho80-1", 1:7, lwd=1.5, ylim=c(0.06,1.2), title="pho80")
plainMeanTrace(x, "pho80.scPho4-1", 1:7, lwd=1.5, ylim=c(0.06,1.2), title="pho80.scPho4")
plainMeanTrace(x, "pho4", 1:7, lwd=1.5, ylim=c(0.06,1.2), title="pho4")
```

### 3. Estimate growth rate from the curve
In this section, I'd like to compute two statistics for each growth curve
(1) Growth rate during the exponential stage, and
(2) Final biomass yield

For #1, the procedures are as follows
* Determine the linear range (there should be little lag time)
* Linear fit on log2(OD) ~ Time(hr)
* Extract the doubling time by calculating 1/coef(fit)[2]
* An alternative method is to calculate the derivative at discreet time points and take the maximum rate

For #2, I will measure the rate at 20hr time point
```{r Measure growth rates, echo=FALSE, fig.width=12, fig.height=18, cache=TRUE}
# Prepare data   
# a===============a
# parameters for calculating. vary them as needed
range <- c(2:20)
t.yield <- 20 # sample at 20hr point
i.yield <- which( Time==t.yield )
# calculate yield
yield <- log2( data[i.yield,] / data[2,] )
# define subfunction for apply
calc <- function(x,range,mode=2) {
  ex <- data.frame(od=x[range], t=Time[range])
  tmp.max.rate <- growthRate(ex, mode, plot=FALSE)
  return(tmp.max.rate)
}
# calculate max.rate using growthRate(..., mode=2)
max.rate <- apply(data,2,calc,range)
# calculate the mean / sd of the max.rate and yield
meanRate <- myMeanSD(rbind(max.rate),design.mat)
meanYield <- myMeanSD(rbind(yield),design.mat)

# Plot data  
# a==============a
# barplot, limit to the first 4 conditions
layout(matrix(c(1,2,3),nrow=3))
par(cex.axis=1.5,cex.lab=1.5,cex.main=1.8)->oldpar
# 0. prepare data
strain.remove <- which(unique(meanRate$design.mat$strain) == "empty")
cond.use <- 1:7
# 1. growth rate
rate.plot <- matrix(meanRate$mean, nrow=length(ST))[-strain.remove,cond.use]
rate.plot[1:3,1] <- rep(NA,3)
rate.sd <- matrix(meanRate$sd, nrow=length(ST))[-strain.remove,cond.use]
rownames(rate.plot) <- as.character(unique(meanRate$design.mat$strain)[-strain.remove])
colnames(rate.plot) <- unique( paste(meanRate$design.mat[,"Glucose"], 
                                     meanRate$design.mat[,"Pi"]) )[cond.use]
colors <- c(gray(.5),"orange2","green3","blue3")[c(1,2,2,3,4,4)]
pos <- barplot(rate.plot,beside=TRUE,col=colors,ylim=c(40,160), xpd=FALSE, xlab="Condition = Glucose + Phosphate", ylab="min", main="Minimum doubling time", names.arg=NULL)
arrows( pos, rate.plot, pos, rate.plot+rate.sd, angle=90, length=.08, lwd=.8)
legend("topleft",legend=rownames(rate.plot), fill=colors, cex=1.5)
abline(h=seq(40,160,20),col=gray(.7), lwd=.6)

# 2. yield at 20hr
yield.plot <- matrix(meanYield$mean, nrow=length(ST))[-strain.remove,cond.use]
yield.sd <- matrix(meanYield$sd, nrow=length(ST))[-strain.remove,cond.use]
rownames(yield.plot) <- as.character(unique(meanYield$design.mat$strain)[-strain.remove])
colnames(yield.plot) <- unique( paste(meanYield$design.mat[,"Glucose"], 
                                      meanYield$design.mat[,"Pi"]) )[cond.use]
pos <- barplot(yield.plot,beside=TRUE,col=colors, xpd=FALSE, ylim=c(0,4), 
               xlab="Condition = Glucose + Phosphate", ylab="log2 OD(20hr) / OD(init)",
               main="Yield relative to T0 at 20hr") 
arrows( pos, yield.plot, pos, yield.plot+yield.sd, angle=90, length=.08, lwd=.8)

# 3. log fit curve
start.t <- 6 # start from 1.5 hr measurement
end.rng <- 12:20 # try several different range
log.fit.rate <- numeric(ncol(data))
log.fit.rate <- apply(data, 2, function(x) {
  arr.log.fit <- sapply(end.rng, function(j) calc(x, start.t:j,mode=1))
  return(min(arr.log.fit))
  } )
# special treatment for No Pi
noPi <- apply(data[,61:80], 2, function(x) {
  arr.log.fit <- sapply(8:12, function(j) calc(x, 2:j,mode=1))
  return(min(arr.log.fit))
  } )
log.fit.rate[61:80] <- noPi
meanLogRate = myMeanSD(rbind(log.fit.rate), design.mat)
# plot
log.rate.plot <- matrix(meanLogRate$mean, nrow=length(ST))[-strain.remove,cond.use]
log.rate.sd <- matrix(meanLogRate$sd, nrow=length(ST))[-strain.remove,cond.use]
rownames(log.rate.plot) <- as.character(unique(meanRate$design.mat$strain)[-strain.remove])
colnames(log.rate.plot) <- unique( paste(meanLogRate$design.mat[,"Glucose"], 
                                         meanLogRate$design.mat[,"Pi"]) )[cond.use]
# remove the first three values in the 1st column
log.rate.plot[1:3,1] <- rep(NA,3)
pos <- barplot(log.rate.plot,beside=TRUE,col=colors, ylim=c(60,200), xpd=FALSE, xlab="Condition = Glucose + Phosphate", ylab="min", main="Log linear fit doubling time", names.arg=NULL)
arrows( pos, log.rate.plot, pos, log.rate.plot+log.rate.sd, angle=90, length=.08, lwd=.8)
abline(h=seq(60,200,20),col=gray(.7), lwd=.6 )
```
#### Example: 