RNASeq_analysis | data: 11/13/2013
========================================================
This script takes in raw counts and determine Differential Expression (DE) 
for every gene of interest.
Data-wise, it combines the 06/15/2013 and 11/13/2013 data
hebin
18 nov 2013
Read data
------------------
```{r Read_Data, message=FALSE, echo=FALSE, cache=TRUE}
library(limma)  # note: here the loading order is important. If limma is loaded last, it will mask the plotMA function
library(DESeq2) #       from DESeq2
setwd("~/Documents/work/Pho/Result/RNA-seq/Main")
# Scer
sc.sense <- read.csv("ScerRNAseq_sense.csv",row.names=1)
# sc.anti <- read.csv("ScerRNAseq_anti.csv",row.names=1)
sc.info <- read.table("ScerRNAseq_sample_info.txt",col.names=c("Sample","Genotype", "Size"),as.is=TRUE)
sc.genes <- rownames( sc.sense )

# Cgla
cg.sense <- read.csv("CglaRNAseq_sense.csv",row.names=1)
# cg.anti <- read.csv("CglaRNAseq_anti.csv",row.names=1)
cg.info <- read.table("CglaRNAseq_sample_info.txt",col.names=c("Sample","Genotype","Size"),as.is=TRUE)
cg.genes <- rownames( cg.sense )

# misc annotation stuff
# convert Cgla genes to orthologs in Scer, based on orthogroup assignment from Aviv's database
# note that for genes with more than one potential orthologs, the first one in the list is used
cg.sc.ortholog <- read.table("Cgla-Scer-orthologs.txt",na.string="NONE",fill=TRUE,as.is=TRUE)
cgToSc <- cg.sc.ortholog$V2; names(cgToSc) <- cg.sc.ortholog$V1

sgdName <- read.table("sgdToName.txt",as.is=TRUE)
sgdToName <- sgdName$V2; names(sgdToName) <- sgdName$V1
nameToSgd <- sgdName$V1; names(nameToSgd) <- sgdName$V2

xu.genes <- c("PHO8","VTC3","SPL2","PHO81","PHO86","PHO5","PHO11","PHO84","PHO89","PHM6","PHO12","GIT1","VTC1","VTC4","CTF19","YAR070C","ENA2","VTC2","ENA1","GDE1","PHM8","VIP1","CBF1","DDP1","YJL119C","HOR2","YNL217W")
Xu.genes <- nameToSgd[ xu.genes ]
all.80 <-c("YAL005C","YAR068W","YAR070C","YAR071W","YBR072W","YBR093C","YBR157C","YBR169C","YBR296C","YCL040W","YCR098C","YDL204W","YDR005C","YDR019C","YDR039C","YDR040C","YDR270W","YDR281C","YDR481C","YDR516C","YEL011W","YEL065W","YER037W","YER062C","YER072W","YFL004W","YFR053C","YGR233C","YHL035C","YHL040C","YHL047C","YHR136C","YHR138C","YHR214W-A","YHR215W","YIL074C","YIL169C","YJL012C","YJL117W","YJL119C","YJR060W","YKL001C","YKR034W","YKR080W","YLL026W","YLR109W","YLR136C","YLR142W","YLR214W","YLR303W","YLR327C","YLR410W","YLR438W","YML123C","YML128C","YMR011W","YMR058W","YMR105C","YMR173W","YMR173W-A","YMR195W","YMR251W","YMR251W-A","YNL217W","YNR069C","YOL086C","YOL155C","YOL158C","YOR163W","YOR173W","YOR344C","YOR347C","YOR382W","YOR383C","YPL018W","YPL019C","YPL054W","YPL110C","YPR167C")
```

Use DESeq
------------------
### 1. Subfunctions for calculating pairwiseDE
```{r Subfunctions_4_DESeq, message=FALSE}
# Function for calculating DE genes given the data, design 
PairwiseDE <- function( countTable, design, cond.A, cond.B ){
  index.A <- which( design==cond.A )
  index.B <- which( design==cond.B )
  subTable <- countTable[, c(index.A,index.B)]
  conditions <- factor( c(rep(cond.A,length(index.A)), rep(cond.B, length(index.B))) )
  cds <- newCountDataSet( subTable, conditions )
  cds <- estimateSizeFactors( cds )
  cds <- estimateDispersions( cds )
  res = nbinomTest( cds, cond.A, cond.B )
  res1 <- res[ !is.na(res$padj), ]
  return( res )
}

# Function to report significant genes at a given fdr and log2FoldChange threshold
sigGene <- function( res, fdr=0.05, log2FoldChange=1, direction="up", baseMean=0 ) {
  par <- paste("sigGene --fdr=", fdr, " --foldchange=", 2^log2FoldChange, " --direction=", direction, " --baseMean=",baseMean, sep="", collapse="")
  if( direction != "up" & direction != "down" ){
    list = rep(FALSE, nrow(res))
  }
  else if( direction == "up"){
    list = res$padj < fdr & res$log2FoldChange > log2FoldChange & res$baseMean > baseMean
  }
  else if( direction == "down"){
    list = res$padj < fdr & res$log2FoldChange < -log2FoldChange & res$baseMean > baseMean
  }
  list = replace(list, is.na(list), FALSE)
  attr(list, "par") <- par
  cat( par )
  return( list )
}
# Use DESeq2 to call significantly DE genes
PairwiseDE2 <- function( countTable, design, cond.A, cond.B ){
  index.A <- which( design==cond.A )
  index.B <- which( design==cond.B )
  subTable <- countTable[, c(index.A,index.B)]
  conditions <- factor( c(rep(cond.A,length(index.A)), rep(cond.B, length(index.B))), levels=c(cond.A,cond.B) )
  colData <- DataFrame( condition = conditions )
  dds <- DESeqDataSetFromMatrix( countData = subTable, colData = colData, design = ~condition)
  dds <- DESeq( dds )
  return( dds )
}
# Compare two lists
compList <- function(listA, listB, total=0){
  res <- list(common = intersect(listA, listB), A.only = setdiff(listA, listB), B.only = setdiff(listB, listA))
  len <- sapply(res, length)
  cat(paste(c("In both"," A only"," B only"), len, sep=":", collapse="\n"))
  if( total ) { # if total # of genes is given, calculate the enrichment factor and p-value from hypergeometric dist.
    expect <- length(listB) / total * length(listA)
    enrich <- round( len["common"] / expect, 1 )
    p.hypergeom <- phyper(q=len["common"],m=length(listB),n=total-length(listB),k=length(listA),lower.tail=F)
    cat("\n\n")
    cat( sprintf( "Enrichment factor: %.1f\np< %.2g", enrich, p.hypergeom ) )
  }
  return(res)
}
```

### 2. Actual analysis of the data
#### 2.1 Compare scPho4 targets in _pho80_ vs. under starvation in Scer
_Question: what genes change their expression in pho80 bg vs. those under starvation? Which of them are PHO genes, i.e. dependent on Pho4?_
```{r DESeq_in_Scer, cache=TRUE, message=FALSE, echo=FALSE }
baseTh <- 100
# naming conventions: species.pho4.comment.object

# Identify genes in pho80
cat("Identify genes in pho80_del in Scer...")
dds <- list( sc.sc4.no80 = PairwiseDE2( sc.sense, sc.info$Genotype, "wt.highPi", "pho80" ) )
res <- list( sc.sc4.no80 = results(dds[[1]]) )
sig <- list( sc.sc4.no80 = sigGene(res$sc.sc4.no80,baseMean=baseTh))
sigd <- list( sc.sc4.no80 = sigGene(res$sc.sc4.no80,direction="down",baseMean=baseTh))
gene <- list( sc.sc4.no80 = sc.genes[ sig$sc.sc4.no80 ] )

# Identify genes under starvation
cat("Identify genes under starvation in Scer...")
dds$sc.sc4.noPi <- PairwiseDE2( sc.sense, sc.info$Genotype, "wt.highPi", "wt.lowPi")
res$sc.sc4.noPi <- results(dds[[2]])
sig$sc.sc4.noPi <- sigGene(res[[2]], baseMean=baseTh)
sigd$sc.sc4.noPi <- sigGene(res[[2]], direction="down", baseMean=baseTh)
gene$sc.sc4.noPi = sc.genes[ sig$sc.sc4.noPi ]

# Identify genes dependent on Pho4
cat("Identify genes dependent on Pho4 in pho80- bg in Scer...")
dds$sc.sc4.no4 <- PairwiseDE2( sc.sense, sc.info$Genotype, "pho80.pho4", "pho80")
res$sc.sc4.no4 <- results(dds[[3]])
sig$sc.sc4.no4 <- sigGene(res[[3]], baseMean=baseTh)
sigd$sc.sc4.no4 <- sigGene(res[[3]], direction="down", baseMean=baseTh)
gene$sc.sc4.no4 = sc.genes[ sig$sc.sc4.no4 ]

# Identify genes dependent on Pho4 by comparing gene induction in pho80 with or without Pho4
cat("Identify genes dependent on Pho4 in pho80- bg in Scer...")
dds$sc.sc4.no80no4 <- PairwiseDE2( sc.sense, sc.info$Genotype,"wt.highPi", "pho80.pho4")
res$sc.sc4.no80no4 <- results(dds[[4]])
sig$sc.sc4.no80no4 <- sigGene(res[[4]], baseMean=baseTh)
sigd$sc.sc4.no80no4 <- sigGene(res[[4]], direction="down", baseMean=baseTh)
gene$sc.sc4.no80no4 = sc.genes[ sig$sc.sc4.no80no4 ]
```

```{r plotMA_Scer, echo=FALSE, fig.width=10, fig.height=12 }
# compare pho80 and starvation
comp <- list( sc.sc4.no80_noPi = table( "pho80"=sig[[1]], "noPi"=sig[[2]] ) )
cat("Compare pho80- and Pi starvation genes")
print( addmargins(comp[[1]]) ) # print overlaps
layout(matrix(1:4,ncol=2)) # MA plots.
DESeq2::plotMA( dds$sc.sc4.no80, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 / wt.highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$sc.sc4.noPi, pvalCutoff=.01, ylim=c(-6.64,6.64), main="wt noPi / highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$sc.sc4.no80no4, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 pho4 / wt.highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$sc.sc4.no4, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 / pho80 pho4");abline(v=100,h=1,lty=2)
```

```{r Scer_compare_gene_set_overlap}
cat( "Genes defined by Xu and found in starvation conditions ...")
tmp <- compList( gene$sc.sc4.no80, gene$sc.sc4.noPi, 6603 )
gene.list <- cbind( sig$sc.sc4.no80, sig$sc.sc4.noPi, sig$sc.sc4.no80 & !sig$sc.sc4.no80no4, FALSE )
rownames(gene.list) <- sc.genes; colnames(gene.list) <- c("pho80", "no Pi", "require Pho4", "Xu Pho4-genes")
gene.list[ all.80, 4 ] <- TRUE
vennCounts <- vennCounts(gene.list[,-1])
vennDiagram( vennCounts, circle.col=c("black","blue","green") )
```

#### 2.2 Compare cgPho4 targets in _pho80_ vs. under starvation in Cgla
```{r DESeq_in_Cgla, cache=TRUE, message=FALSE, echo=FALSE }
baseTh <- 100
# naming conventions: species.pho4.comment.object

# Identify genes in pho80
cat("Identify genes in pho80_del in Cgla...")
dds$cg.cg4.no80 = PairwiseDE2( cg.sense, cg.info$Genotype, "wt.highPi", "pho80" ) 
res$cg.cg4.no80 = results(dds[[1]]) 
sig$cg.cg4.no80 = sigGene(res$cg.cg4.no80,baseMean=baseTh)
sigd$cg.cg4.no80 = sigGene(res$cg.cg4.no80,direction="down",baseMean=baseTh)
gene$cg.cg4.no80 = cg.genes[ sig$cg.cg4.no80 ] 

# Identify genes under starvation
cat("Identify genes under starvation in Cgla...")
dds$cg.cg4.noPi <- PairwiseDE2( cg.sense, cg.info$Genotype, "wt.highPi", "wt.lowPi")
res$cg.cg4.noPi <- results(dds[[2]])
sig$cg.cg4.noPi <- sigGene(res[[2]], baseMean=baseTh)
sigd$cg.cg4.noPi <- sigGene(res[[2]], direction="down", baseMean=baseTh)
gene$cg.cg4.noPi = cg.genes[ sig$cg.cg4.noPi ]

# Identify genes dependent on Pho4
cat("Identify genes dependent on Pho4 in pho80- bg in cgla...")
dds$cg.cg4.no4 <- PairwiseDE2( cg.sense, cg.info$Genotype, "pho80.pho4", "pho80")
res$cg.cg4.no4 <- results(dds[[3]])
sig$cg.cg4.no4 <- sigGene(res[[3]], baseMean=baseTh)
sigd$cg.cg4.no4 <- sigGene(res[[3]], direction="down", baseMean=baseTh)
gene$cg.cg4.no4 = cg.genes[ sig$cg.cg4.no4 ]

# Identify genes dependent on Pho4 by comparing gene induction in pho80 with or without Pho4
cat("Identify genes dependent on Pho4 in pho80- bg in cgla...")
dds$cg.cg4.no80no4 <- PairwiseDE2( cg.sense, cg.info$Genotype,"wt.highPi", "pho80.pho4")
res$cg.cg4.no80no4 <- results(dds[[4]])
sig$cg.cg4.no80no4 <- sigGene(res[[4]], baseMean=baseTh)
sigd$cg.cg4.no80no4 <- sigGene(res[[4]], direction="down", baseMean=baseTh)
gene$cg.cg4.no80no4 = cg.genes[ sig$cg.cg4.no80no4 ]
```

```{r plotMA_Cgla, echo=FALSE, fig.width=10, fig.height=12 }
# compare pho80 and starvation
comp$cg.cg4.no80_noPi = table( "pho80"=sig[[1]], "noPi"=sig[[2]] ) 
cat("Compare pho80- and Pi starvation genes")
print( addmargins(comp[[1]]) ) # print overlaps
layout(matrix(1:4,ncol=2)) # MA plots.
DESeq2::plotMA( dds$cg.cg4.no80, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 / wt.highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$cg.cg4.noPi, pvalCutoff=.01, ylim=c(-6.64,6.64), main="wt noPi / highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$cg.cg4.no80no4, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 pho4 / wt.highPi");abline(v=100,h=1,lty=2)
DESeq2::plotMA( dds$cg.cg4.no4, pvalCutoff=.01, ylim=c(-6.64,6.64), main="pho80 / pho80 pho4");abline(v=100,h=1,lty=2)
```

